A Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-02 02-ceu.tex
Revisão 2.1

\CEU defines time as a discrete sequence of reactions to unique external
input events.
%
These external input events are received from the environment; each of them
delimits a new logical unit of time and triggers a corresponding reaction.
%
The life-cycle of a program in \CEU can be summarized as
follows~\cite{ceu.sensys13}.
%
\begin{enumerate}
\item The program initiates the ``boot reaction'' in a single trail
  (parallel constructs may create new trails).
  %
\item Active trails execute until they await or terminate, one after
  another.  This step is called a \emph{reaction chain}, and always runs in
  bounded time.
  %
\item When all its trails are blocked, the program goes idle and the
  environment takes control.
  %
\item On the occurrence of a new external input event, the environment
  awakes \emph{all} trails awaiting that event, and the program goes back to
  step~(ii).
\end{enumerate}

A program must react to an event completely before handling the next one.
%
By the synchronous hypothesis, the time the program spends in step~(ii)
above is conceptually zero (in practice, negligible).
%
Thus, from the point of view of the environment, the program is always
idle---on step~(iii).
%
In practice, if a new external input event occurs while a reaction is being
computed, the event is saved on a queue, which effectively schedules it to
be processed by a subsequent reaction.

@@

The processing of external input events induces a discrete notion of time in
\CEU.
%
Figure~\ref{fig.ticks} illustrates this notion.
%
The continuous timeline shows an absolute reference clock with ``physical
timestamps'' for the event occurrences (e.g., event~\code{C} occurs at
$17ms521us$).
%
The discrete timeline shows how the same occurring events fit in the logical
notion of time of \CEU.
%
The boot reaction \code{boot-0} happens before any input on program startup.
%
Event~\code{A} ``physically'' occurs during \code{boot-0} but, because time
is discrete, its corresponding reaction can only execute afterwards, at
logical instant~\code{A-1}.
%
Similarly, event~\code{B} occurs during~\code{A-1} and its reaction is
postponed to execute at~\code{B-2}.
%
Event~\code{C} also occurs during~\code{A-1} but its reaction must also wait
for~\code{B-2} to execute and so it is postponed to execute at~\code{C-3}.
%
Event~\code{D} occurs during an idle period and can start immediately at
\code{D-4}.
%
Finally, two instances of event~\code{E} occur during~\code{D-4}; they are
handled in the subsequent reactions~\code{E-5} and~\code{E-6}.

@@

In practice, the synchronous hypothesis for \CEU holds if reaction rate is
faster than the rate of incoming input events.
%
Otherwise, the program continuously accumulates a delay between the real
occurrence time and actual reaction time of events.
%
In the soft real-time systems targeted by \CEU (e.g., sensor networks,
multimedia systems, interactive games, etc.) such delay and postponed
reactions might be tolerated by users as long as they are infrequent and the
application does not take too long to catch up with real time.

>>>>

\CEU defines time as a discrete sequence of reactions to unique external
input events.
%
External input events are received from the environment, and each delimits a
new logical unit of time that triggers an associated reaction.
%
The life-cycle of a program in \CEU can be summarized as
follows~\cite{ceu.sensys13}:
%
\begin{enumerate}[i]
\item The program initiates a ``boot reaction'' in a single trail (parallel
      constructs may create new trails).
\item Active trails execute until they await or terminate, one after
      another.  This step is called a \emph{reaction chain}, and always runs in
      bounded time.
\item When all trails are blocked, the program goes idle and the environment
      takes control.
\item On the occurrence of a new external input event, the environment
      awakes \emph{all} trails awaiting that event, and the program goes back to
      step~(i).
\end{enumerate}

A program must react to an event completely before handling the next one.
%
By the synchronous hypothesis, the time the program spends in step~(ii) is
conceptually zero (in practice, negligible).
%
Hence, from the point of view of the environment, the program is always
idle on step~(iii).
%
In practice, if a new external input event occurs while a reaction executes,
the event is saved on a queue, which effectively schedules it to be processed
in a subsequent reaction.

@@

The sequential processing of external input events induces a discrete notion of
time in \CEU, as illustrated in Figure~\ref{fig.ticks}.
%
The continuous timeline shows an absolute reference clock with ``physical
timestamps'' for the event occurrences (e.g., event~\code{C} occurs at
$17ms521us$).
%
The discrete timeline shows how the same occurring events fit in the logical
notion of time of \CEU.
%
The boot reaction \code{boot-0} happens before any input, at program startup.
%
Event~\code{A} ``physically'' occurs during \code{boot-0} but, because time
is discrete, its corresponding reaction only executes afterwards, at logical
instant~\code{A-1}.
%
Similarly, event~\code{B} occurs during~\code{A-1} and its reaction is
postponed to execute at~\code{B-2}.
%
Event~\code{C} also occurs during~\code{A-1} but its reaction must also wait
for~\code{B-2} to execute and so it is postponed to execute at~\code{C-3}.
%
Event~\code{D} occurs during an idle period and can start immediately at
\code{D-4}.
%
Finally, two instances of event~\code{E} occur during~\code{D-4}; they are
handled in the subsequent reactions~\code{E-5} and~\code{E-6}.

@@

In practice, the synchronous hypothesis for \CEU holds if reactions execute
faster than the rate of incoming input events.
%
Otherwise, the program would continuously accumulate delays between physical
occurrences and actual reactions for the input events.
%
In the soft real-time systems targeted by \CEU (e.g., sensor networks,
multimedia systems, interactive games, etc.) such delay and postponed
reactions might be tolerated by users as long as they are infrequent and the
application does not take too long to catch up with real time.
%
Note that the synchronous semantics is the norm in typical event-driven
systems, such as event dispatching in UI toolkits, game loops in game engines,
and clock ticks in embedded systems.
